Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-04 submission.tex
mindre endringer i introduksjonen

\section{Introduction}
\label{sec:intro}
PoS tagging is an important preprocessing step for many NLP tasks,
such as dependency parsing \cite{Niv:Hal:Kub:07,Haj:Cia:Joh:09}, named
entity recognition \cite{Tjo:DeM:03} and sentiment analysis
\cite{Wil:Wie:Hof:09}. Whereas much effort has gone into the
development of PoS-taggers, to the effect that this task is often
considered more or less a solved task, considerably less effort has
been devoted to the empirical evaluation of the PoS tag sets
themselves. Error analysis of PoS taggers indicate that, whereas tagger
improvement though means of learning algorithm or feature engineering seems to
have reached something of a plateau, linguistic assessment of the distinctions
made by the PoS tag set may be a venue worth investigating further
\cite{Man:11}. Clearly, the utility of a PoS tag set is tightly coupled with
the downstream task for which it is performed. Even so, PoS tag sets are
usually employed in a ``one size fits all'' fashion, regardless of the
requirements posed by the task which makes use of this information.

It is well known that syntactic parsing often benefits from quite
fine-grained morphological distinctions
\cite{Zha:Niv:11,See:Kuh:13}. Morphology interacts with syntax though
phenomena such as agreement and case marking and incorporating
information on morphological properties of words can therefore often
improved parsing performance. However, in a realistic setting where
the aim is to automatically parse raw text, the generation of
morphological information will often require a separate step of
morphological analysis that can be quite costly.

In this paper, we optimize a Norwegian PoS tag set for the task of
data-driven dependency parsing.  We show that the introduction of
morphological distinctions not present in the original tag set, whilst
compromising tagger accuracy, actually leads to significantly improved
parsing accuracy.  In a set of experiments, various morphological
distinctions are introduced and evaluated both intrinsically, i.e. in
terms of PoS tagging accuracy, and extrinsically, in terms of parsing
accuracy.  Our results show that the introduction of morphological
distinctions not present in the original tag set, whilst compromising
tagger accuracy, actually leads to significantly improved parsing
accuracy. This optimization allows us to bypass the additional step of
morphological analysis, framing the whole preprocessing problem as a
simple tagging task.

The article is structured as follows. We start out by reviewing previous work
on tag set evaluation in Section \ref{sec:prev} and Section \ref{sec:data}
provides an overview of the treebank used for experimentation. Section
\ref{sec:setup} describes the experimental setup used in our work and Section
\ref{sec:optimization} goes on to provide the results from our optimization
experiments. Finally, Section \ref{sec:summary} summarizes our main findings
and discusses some avenues for future work.
The article is structured as follows. We start out by reviewing
previous work on tag set evaluation in Section \ref{sec:prev} and
Section \ref{sec:data} provides an overview of the treebank used for
experimentation. Section \ref{sec:setup} describes the experimental
setup used in our work and Section \ref{sec:optimization} goes on to
provide the results from our optimization experiments. Finally,
Section \ref{sec:summary} summarizes our main findings and discusses
some avenues for future work.

>>>>

\section{Introduction}
\label{sec:intro}
Part-of-speech (PoS) tagging is an important preprocessing step for many NLP tasks,
such as dependency parsing \cite{Niv:Hal:Kub:07,Haj:Cia:Joh:09}, named
entity recognition \cite{Tjo:DeM:03} and sentiment analysis
\cite{Wil:Wie:Hof:09}. Whereas much effort has gone into the
development of PoS-taggers -- to the effect that this task is often
considered more or less a solved task -- considerably less effort has
been devoted to the empirical evaluation of the PoS tag sets
themselves. Error analysis of PoS-taggers indicate that, whereas
tagging improvement through means of learning algorithm or feature
engineering seems to have reached something of a plateau, linguistic
and empirical assessment of the distinctions made in the PoS tag set may be a venue
worth investigating further \cite{Man:11}.  Clearly, the utility of a
PoS tag set is tightly coupled with the the downstream task for which
it is performed. Even so, PoS tag sets are usually employed in a ``one
size fits all'' fashion, regardless of the requirements posed by the
task which makes use of this information.

It is well known that syntactic parsing often benefits from quite
fine-grained morphological distinctions
\cite{Zha:Niv:11,See:Kuh:13}. Morphology interacts with syntax though
phenomena such as agreement and case marking, and incorporating
information on morphological properties of words can therefore often
improved parsing performance. However, in a realistic setting where
the aim is to automatically parse raw text, the generation of
morphological information will often require a separate step of
morphological analysis that can be quite costly.

In this paper, we optimize a PoS tag set for the task of
data-driven dependency parsing of Norwegian. We report on a set of experiments where various morphological
distinctions are introduced to the PoS annotations and evaluated both intrinsically, i.e. in
terms of PoS-tagging accuracy, and extrinsically, in terms of parsing
accuracy.  Our results show that the introduction of morphological
distinctions not present in the original tag set, whilst compromising
tagger accuracy, actually leads to significantly improved parsing
accuracy. This optimization allows us to bypass the additional step of
morphological analysis, framing the whole preprocessing problem as a
simple tagging task.

The article is structured as follows. We start out by reviewing
previous work on tag set evaluation in Section \ref{sec:prev}, while
Section \ref{sec:data} details the treebank that provides the initial gold 
annotations used for our experiments. Section \ref{sec:setup} describes the experimental
setup used in our work and Section \ref{sec:optimization} goes on to
provide the results from our tag set optimization. Finally,
Section \ref{sec:summary} summarizes our main findings and discusses
some avenues for future work.
